Technologies of Semantic WEB as an environment of application development and integration

### 2022-02-04 paper-aicts-cherkashin.tex
Sorting references
Release

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=true,           % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=false,        % show Acrobat’s toolbar?
    pdfmenubar=false,        % show Acrobat’s menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={},    % title
    pdfauthor={Evgeny Cherkashin},     % author
    pdfsubject={Technologies of Semantic WEB as an environment of application development and integration},   % subject of the document
    pdfnewwindow=true,      % links in new PDF window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=green,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=blue           % color of external links
  }

\usepackage{pifont}
\usepackage{xcolor}
\graphicspath{{pics/}}

@@

%%
%% This command is for the conference information
\conference{2nd International Workshop on Advanced Information and Computation Technologies and Systems, December XX--XX, 2021, Irkutsk, Russia}

@@

%%
%% The abstract is a summary of the work to be presented in the
%% article.
\begin{abstract}
  Semantic Web technologies and their standards are becoming productive assets for software development and integration on various design and implementation levels.  These include OSI level 6 data representation, database and file storage formats, the source data for text indexes, document authoring, hypertext markup engines, a representation for abstract software models, \emph{etc.}  They are also smoothly combined with the development and data analysis tools, giving rise to a new common-ground environment for integrating distributed applications on a semantic level.

@@

In Russia, domain model designers generally prefer to develop local vocabularies instead of spending time for the routine work of Internet surfing for existing conceptual models and their enrichment.  It is due to underdevelopment of the metadata resources (metavocabularies) due to lack of a domain classification system as compared, \emph{e.g.} to Universal Decimal Classification (UDC) used in the bibliography.  Vocabularies also can have common parts, describing the same or similar entities with different terms and relations, \emph{e.g.} semantically more adapted to a specific problem-solving.

@@

Each entity described in SW has its unique global IRI, its virtual or real internet address.  Document entities' resource IRIs are formed accounting the document structure, \emph{e.g.} a hierarchy of contexts.  Using IRIs in applications solves the fundamental problem of cross-domain and cross-application object identification and forms a basis for a standardized data processing environment.  Entities of various abstract levels can be designated an IRI, including vocabulary definitions (\emph{namespaces}), data types, data containers (files, databases).

@@

KGs and subgraphs are programmatically processed using libraries developed for all papular programming environments.  KG is represented with a triple or quadruple set object.  The set is traversed, triples are filtered according to various conditions.  Filters are usually defined as patterns, where each component is either a free variable/placeholder or a concrete value (IRI, literal), which must match exactly.   Some libraries allow one to execute \texttt{SPARQL} queries locally.  KG is evolved by adding new triples and removing irrelevant ones.  The resulting modified KG is stored as files and published on a server.  An example of a federated SPARQL query is presented in Figure~\ref{fig:sparql-ex1}, which results in counting parts classes as the first level of a three widget.  Each class is named in Russian if \texttt{DBPedia} contains a corresponding label with \texttt{ru} language tag.  Otherwise, the label is figured out of IRI.  This query is initially run at a local server, and the server initiates a subquery to \texttt{DBPedia.org} endpoint.

@@

In order to present the general architecture of application development, we have to speak about KG definitions and useful properties briefly of KGs.
% \subsection{Knowledge graph properties as application basis}
% \label{sec:kg-as-basis}
KGs changed the viewpoint to the knowledge base as being a part of the whole totality of knowledge, implying the obeying the global standards and techniques of its acquisition and processing \cite{hogan}.  The following outlines their definitions \cite{hogan}.
 \begin{itemize}
  \item Notions \emph{data} and \emph{knowledge} are converged into ``\emph{something} \emph{known}'';
  \item KG contains data and knowledge in the form of resources and relations, including metadata;
  \item The same KG is being filled in and processed by different independent modules, \emph{e.g.}, withing SPARQL queries with UPDATE statements;
  \item KG fundamentals allow a software developer to \emph{postpone} the formal definition of a data schema;
  \item There are three points of view to graph schemata interpretation:
    \begin{itemize}
    \item \emph{semantic}, which is aimed at domain modeling paying attention to generalization relations (classes);
    \item \emph{validating}, reflecting semantic verification, checking completeness with respect to sets of predefined general relations, and
    \item \emph{emergent}, observed or deduced from exploitation experience analysis, as a set of new generalized structures and \emph{refactoring} of the KG.
    \end{itemize}
  \end{itemize}

@@

With general constraints implied by KG based domain modeling, the design of application modules are to be built on standardized data interchange formats, which are to be easily converted from and to KG representation, as well as applications are to be constructed out of service modules supporting SW endpoints (SPARQL) whenever possible.

@@

Our R\&D resulted in a set of components and a general application architecture showed in Figure~\ref{fig:arch-services}.
\begin{figure}[bth]
  \begin{center}
    \includegraphics[width=0.9\linewidth]{architecture-mda-lod-ext-general.pdf}
  \end{center}
  \begin{center}
      \textbf{Abbreviations}%\\[1ex]
    \end{center}
    \begin{quote}\scriptsize
      % T-Module is Transformation module,
      MDA is Model-Driven Architecture,
      CIM is Computationally Independent Model,
      PIM is Platform Independent Model,
      PSM is Platform Specific Model,
      T-Box is Terminological Box,
      KB is Knowledge Base,
      % A-Box is Instance Box,
      % NGS is Next-Generation Sequencing,
      DB is Database.
    \end{quote}
  \caption{Architecture of application instrumental environment as a set of components}
  \label{fig:arch-services}
\end{figure}
The kernel module of applications is \texttt{KG server} (B2) built on the base of ClioPatria \cite{cliopatria} with web-server and Pengines plug-ins.  ClioPatria is a web-application implemented in SWI-Prolog \cite{swi} being ISO-compliant Prolog implementation.  KG functions are implemented with SW package, supplied in standard SWI-Prolog distribution.  Pengines service enables SW web-applications to utilize logical inference on the server-side \cite{pengines} with browser JavaScript environment integration.  Pengine's knowledge base is extended with the programmer's Prolog modules.  Kernel module stores a common global part of A-~and T-boxes of applications, \emph{i.e.} it is application database.  Other modules provide an application (C3) with services.

Authoring tool (B1) is used to provide document generation, and integration \cite{zont19}.  This tool enables document content and structure inference in the browser utilizing templates and data from other web pages and websites providing LOD\footnote{Abbreviation of Linked Open Data} \cite{lod} capabilities.  Programmer implements a view of an object as a web page with an extended version of RDFa.  The page is generated within any programming runtime, \emph{e.g.} PHP.  The authoring tool engine collects web page chunks and composes the document as an XHTML page.  This composed page can be stored in a KG as a set of triples and indexed in the Text indexing module (C1), printed with the browser's printing engine, or converted to PDF.  The tool allows the user to change document content in WYSIWYG mode.  Its general structure is defined by changing the template on the server-side.

Text indexing engine (C1) provides a service for storing documents represented as KG triples, JSONs, and BLOBS of any format with corresponding text layers.  This module has two implementations.  One is built on the Sphinx Search engine, and the second is Elasticsearch.  Elasticsearch supports JSON as the only document format and is easily integrated with KG using JSON-LD for triple representation.  Sphinx Search is much faster in text and database records indexing and consumes less RAM as it is implemented in C++.  The choice of an engine depends on the primary document representation used in an application.

BLOB stored documents having no markup (PDFs, scanned documents) usually contain valuable data to be revealed for application use.  The documents could be report data or scientific paper content, DJVU files, or raster images.  Data recognition is implemented in the PDF/Text document processing module (C2), where data processing workers query the Text indexing module for BLOB documents and add recognized text and high-order structure layers.  All obtained layers are stored in the database of C1, and data of general interest are to be converted in triples of KG.

@@

The aim of this R\&D \cite{zont19, aicts2020} is to develop software support of the processes of NGS analysis with techniques and software for a visual representation of the computational process for so-called amplicon analysis so that domain specialists would be able to compose computational pipelines, which are executed on distributed heterogeneous computing resources (clouds).  Later, the visual models are used to organize the computational process, data store, and particular NGS data retrieval services.

@@

All transformations were carried out with an object-oriented (OO) Logtalk programming environment representing modules that logically infer new data descriptions of the target objects.  Thanks to OO, the transformation has two levels of abstraction in its representation.  The higher one is a network of objects sending messages via their interfaces.  At the lower one, rules analyze and produce new conclusions on generated object properties.

The PSM objects must have an order in their description as they represent a model of source code to be generated.  The order is organized with blocks of \verb|assert|ed statement and subobjects sequences referencing KG elements.  Blocks can be appended and pretended with new elements.  At the rendering sources stage, the three blocks are traversed in-depth.  It leverages the known \verb|llvmlite| source code generation approach.  Figure~\ref{fig:block-seq} shows the main idea behind PSM object synthesis.
\begin{figure}
  \centering
\begin{multicols}{2}
\begin{minted}[fontsize=\scriptsize]{logtalk}
:- object(class, specializes(code_block),
   imports([named])). % Category of named entities
:- public([classlist/1, methods/1, attributes/1]).
% . . . . . . . . . . . . . .
renderitem(Item, Result):-      % proceed with default
    ^^renderitem(Item, Result). % rendering
render(Result):-         % Source generator
    ^^render(Name),      % implemented in a category
    ( ::item(classlist(List)) ->
     % . . . . . . . . . . .
        [Name]) ),
    ( ::item(attributes(Attributes))->
     % . . . . . . . . . . .
        [DefAttrList]),
      Attributes::items(InstanceAttrs),
      findall(S, ( % initialize attributes
         % . . . . . . . . .
         ), AttrAssigns),
        root::unindent,
        AttrList=[ConstructorDef|AttrAssigns];
         % . . . . . . . . .
        AttrList=[ConstructorDef, Pass] ),
    ( ::item(methods(Methods))-> % If any ...
      Methods::render(MethodList);
      MethodList=[] ),
    lists::append(AttrList,MethodList,StringList),
    root::unindent, Result=[Signature|StringList].
:- end_object.
\end{minted}
      \includegraphics[width=0.7\linewidth]{code_block_class.pdf}
\end{multicols}
  \caption{Procedure of dataflow module source code model synthesis}
  \label{fig:block-seq}
\end{figure}
This example of the application of KG utilizes strongly the property of data representation flexibility for description of the software object model data as well as one central point of data store and access.  The disadvantageous properties of RDF, namely, the weak expressiveness with triples, was neglected by predicate and object encapsulation.

@@

This project \cite{authoring} is aimed at the construction of an environment for document authoring out of user-editable template elements, database query results, LOD-compliant web pages, and other document data.  The target document is a web page where content is inferred while loaded by the browser.

All document elements are either context structures or templates filled with other elements.  A context is an RDF hierarchical representation of a document data structure, providing data to be filled in a current rendered template.  The document is rendered from its main template.  All elements are SW resources.  Templates can contain JavaScript subroutines for implementing special services, for example, table of contents out of CSS \verb|class| labeled subsections of the generated document.

@@

<main lang="ru" resource="#annotation" typeof="oa:Annotation" id="main-doc-cnt">
<div property="oa:hasTarget" resource="#course-work-prog"></div>
<article property="oa:hasBody" typeof="foaf:Document curr:WorkingProgram"
         resource="#course-work-program" id="main-document">
  <div |\GB{taa:content}|="imei:title-page"></div>
  <div |\GB{taa:content}|="imei:neg-UMK"></div>
  <section id="TOC" class="break-after"> <h2>Table of Contents</h2>
    <div id="tableOfContents"></div>
  </section>
  <section id="course-description" resource="#description"
           property="schema:hasPart" typeof="schema:CreativeWork">
    <div property="schema:hasPart" resource="#purpose"
         typeof="dc:Text cnt:ContentAsText" >
      <div property="cnt:chars" datatype="xsd:string">
        <h2 property="dc:title" datatype="xsd:string">
           Aims and objectives of the discipline (module)</h2>
        <p>The aim of teaching the discipline ...</p>
      </div>
   </div>
  . . . . . . . .
\end{minted}
  \caption{Part of a template representation}
  \label{fig:auth-code}
\end{figure}

For document markup (document data representation), the following vocabularies are used:
\begin{itemize}
  \item Friend-of-a-friend (\textbf{foaf}), agent information: individuals, legal entities, program agents;
  \item Provenance (\textbf{prov}), references between documents;
  \item Dublin Core (\textbf{dc}), web page annotation mark up;
  \item DBPedia resource (\textbf{dbr}), references to instant objects and classes;
  \item Schema.org (\textbf{schema}), Google, Yandex, Yahoo, \emph{etc}. searchable objects, structural elements;
  \item The Bibliographic Ontology (\textbf{bibo}), literature reference mark up.
\end{itemize}

@@

GIS represents spatial data in semantic-defined layers.  For each object of a GIS layer, one can associate a set of attribute values of auxiliary data.  The attribute set is the same for each object of a layer, regardless the attribute's value is defined or not.  Empty values are represented as ``\texttt{null}''s.  In the case of geological exploration, when many attributes are undefined, this approach leads to sparsely filled tables.  In this case, structure modification and data analysis algorithms are required to utilize additional data checking stages when using standard relation operations (SELECT, UPDATE, DELETE).

@@

\bibitem{tbl} Tim Berners-Lee, James Hendler, and Ora Lassila.  The semantic web: A new form of web content that is meaningful to computers will unleash a revolution of new possibilities.  Scientific American, Vol.~05, 2001.

\bibitem{hogan} A.~Hogan, E.~Blomqvist, M.~Cochez, C.~D’Amato \emph{et al}. Knowledge Graphs, 2020. \url{https://arxiv.org/abs/2003.02320v5}

\bibitem{lod} C.~Bizer, T.~Heath, T.~Berners-Lee, Linked Data -- The Story So Far.  Int. J. Semantic Web Inf. Syst., 5 (2009), pp.~1--22. \doi{10.4018/jswis.2009081901}

\bibitem{lunina} O.~V.~Lunina.  The digital map of the Pliocene–Quaternary crustal faults in the Southern East Siberia and the adjacent Northern Mongolia.  Geodynamics \& Tectonophysics.  V.~7(3) (2016).  pp.~407-434. (in Russian) \doi{10.5800/GT-2016-7-3-0215}

\bibitem{afs} A.~A.~Gladkov, O.~V.~Lunina. Cartographic service ``Activetectonics''. \url{http://activetectonics.ru/} (access date: 20-Sep-2021)

@@

\bibitem{zont19} E.~Cherkashin, A.~Shigarov, V.~Paramonov. Representation of MDA transformation with logical objects.  International Multi-Conference on Engineering, Computer and Information Sciences (SIBIRCON), Novosibirsk, Russia. (2019) 0913--0918 \doi{10.1109/SIBIRCON48586.2019.8958008}

\bibitem{aicts2020} Evgeniy Cherkashin, Alexey Shigarov. Construction techniques of Baikal microbiome research information-computational environment.  Proceedings for 1st International Workshop on Advanced Information and Computation Technologies and Systems. Irkutsk, Russia, December 7-11, 2020. CEUR-WS.org, online \url{http://ceur-ws.org/Vol-2858/paper2.pdf} (access date: 12-Dec-2021)

\bibitem{gisviewer} E.~Cherkashin, O.~Lunina, L.~Demyanov, A.~Tsygankov. Web-GIS viewer for active faults data represented as a knowledge graph.  Proceedings for 4th Scientific-practical Workshop Information Technologies: Algorithms, Models, Systems. Irkutsk, Russia, September 14, 2021. CEUR-WS.org, online \url{http://ceur-ws.org/Vol-2984/paper8.pdf} (access date: 12-Dec-2021)

\bibitem{cliopatria} J. Wielemaker, W. Beek, M. Hildebrand, J. Ossenbruggen, ``ClioPatria: A
  SWI-Prolog infrastructure for the Semantic Web,'' Semantic Web.
  Vol. 7(5).  P. 529--541 (2016) \doi{10.3233/SW-150191}

\bibitem{shigpaper} Dorodnykh N.O., Yurin A.Y., Shigarov A.O. (2020) Conceptual Model Engineering for Industrial Safety Inspection Based on Spreadsheet Data Analysis.  In: Simian D., Stoica L. (eds) Modelling and Development of Intelligent Systems.  MDIS 2019.  Communications in Computer and Information Science, vol 1126.  Springer, Cham. \doi{10.1007/978-3-030-39237-6_4}

\bibitem{pengines} {Torbj{\"{o}}rn Lager, Jan Wielemaker. Pengines: Web Logic Programming Made Easy.  Theory and Practice of Logic Programming.  Vol.~14, No.~4-5. pp.~539--552. 2014. \doi{10.1017/S1471068414000192}

\bibitem{swi} Jan Wielemaker, Tom Schrijvers, Markus  Triska, Torbj\"o{}rn} Lager. SWI-Prolog. Theory and Practice of Logic Programming, Vol.~12, No.~1-2, pp.~67--96, ISSN~1471-0684.

>>>>

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=true,           % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=false,        % show Acrobat’s toolbar?
    pdfmenubar=false,        % show Acrobat’s menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={},    % title
    pdfauthor={Evgeny Cherkashin},     % author
    pdfsubject={Technologies of Semantic WEB as an environment of application development and integration},   % subject of the document
    pdfnewwindow=true,      % links in new PDF window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=black,          % color of internal links (change box color with linkbordercolor)
    citecolor=black,        % color of links to bibliography
    filecolor=black,      % color of file links
    urlcolor=black           % color of external links
  }
\urlstyle{tt}
\usepackage{pifont}
\usepackage{xcolor}
\graphicspath{{pics/}}

@@

%%
%% This command is for the conference information
\conference{2nd International Workshop on Advanced Information and Computation Technologies and Systems, December 6--10, 2021, Irkutsk, Russia}

@@

%%
%% The abstract is a summary of the work to be presented in the
%% article.
\begin{abstract}
  Semantic Web technologies and their standards are becoming productive assets for software development and integration on various design and implementation levels.  These include OSI level 6 data representation, database and file storage formats, the source data for text indexes, document authoring, hypertext markup engines, a representation for abstract software models, \emph{etc.}  They are also smoothly combined with the development and data analysis tools, giving rise to an environment for integrating distributed applications on a semantic level.

@@

In Russia, domain model designers generally prefer to develop local vocabularies instead of spending time for the routine work of Internet surfing for existing conceptual models and their enrichment.  It is due to underdevelopment of the metadata resources (metavocabularies) due to lack of a domain classification system as compared, \emph{e.g.}, to Universal Decimal Classification (UDC) used in the bibliography.  Vocabularies also can have common parts, describing the same or similar entities with different terms and relations, in particular, semantically more adapted to a specific problem-solving.

@@

Each entity described in SW has its unique global IRI, its virtual or real internet address.  Document entities' resource IRIs are formed accounting the document structure, \emph{e.g.}, a hierarchy of contexts.  Using IRIs in applications solves the fundamental problem of cross-domain and cross-application object identification and forms a basis for a standardized data processing environment.  Entities of various abstract levels can be designated an IRI, including vocabulary definitions (\emph{namespaces}), data types, data containers (files, databases).

@@

KGs and subgraphs are programmatically processed using libraries developed for all papular programming environments.  KG is represented with a triple or quadruple set object.  The set is traversed, triples are filtered according to various conditions.  Filters are usually defined as patterns, where each component is either a free variable/placeholder or a concrete value (IRI, literal), which must match exactly.   Some libraries allow one to execute \texttt{SPARQL} queries locally.  KG is evolved by adding new triples and removing old ones.  The resulting modified KG is stored as files and published on a server.  An example of a federated SPARQL query is presented in Figure~\ref{fig:sparql-ex1}, which results in counting parts classes as the first level of a three widget.  Each class is named in Russian if \texttt{DBPedia} contains a corresponding label with \texttt{ru} language tag.  Otherwise, the label is figured out of IRI.  This query is initially run at a local server, and the server initiates a subquery to \texttt{DBPedia.org} endpoint.

@@

In order to present the general architecture of application development, we have to speak about KG definitions and useful properties briefly of KGs.
% \subsection{Knowledge graph properties as application basis}
% \label{sec:kg-as-basis}
KGs changed the viewpoint to the knowledge base as being a part of the whole totality of knowledge, implying the obeying the global standards and techniques of its acquisition and processing \cite{hogan}.  The following outlines their definitions \cite{hogan}.
 \begin{itemize}
  \item Notions \emph{data} and \emph{knowledge} are converged into ``\emph{something} \emph{known}'';
  \item KG contains data, knowledge and metadata in the form of resources and relations;
  \item The same KG is being filled in and processed by different independent modules, \emph{e.g.}, withing SPARQL queries with UPDATE statements;
  \item KG fundamentals allow a software developer to \emph{postpone} the definition of a data schema;
  \item There are three points of view to graph schemata interpretation:
    \begin{itemize}
    \item \emph{semantic}, which is aimed at domain modeling paying attention to generalization relations (classes);
    \item \emph{validating}, reflecting semantic verification, checking completeness with respect to sets of predefined general relations, and
    \item \emph{emergent}, observed or deduced from exploitation experience analysis, as a set of new generalized structures and \emph{refactoring} of the KG.
    \end{itemize}
  \end{itemize}

@@

With general constraints implied by KG based domain modeling, the design of application modules are to be built on standardized data interchange formats, which are to be easily converted from and to KG representation, as well as applications are to be constructed out of service modules supporting SPARQL endpoints whenever possible.

@@

Our R\&D resulted in a set of components and a general application architecture showed in Figure~\ref{fig:arch-services}.
\begin{figure}[bth]
  \begin{center}
    \includegraphics[width=0.9\linewidth]{architecture-mda-lod-ext-general.pdf}
  \end{center}
  \begin{center}
      \textbf{Abbreviations}%\\[1ex]
    \end{center}
    \begin{quote}\scriptsize
      % T-Module is Transformation module,
      MDA is Model-Driven Architecture,
      CIM is Computationally Independent Model,
      PIM is Platform Independent Model,
      PSM is Platform Specific Model,
      T-Box is Terminological Box,
      KB is Knowledge Base,
      % A-Box is Instance Box,
      % NGS is Next-Generation Sequencing,
      DB is Database.
    \end{quote}
  \caption{Architecture of application instrumental environment as a set of components}
  \label{fig:arch-services}
\end{figure}
The kernel module of applications is KG server (B2) built on the base of ClioPatria \cite{cliopatria} with web-server and Pengines plug-ins.  ClioPatria is a web-application implemented in SWI-Prolog \cite{swi} being ISO-compliant Prolog implementation.  KG functions are implemented with SW package, supplied in standard SWI-Prolog distribution.  Pengines service enables SW web-applications to utilize logical inference on the server-side \cite{pengines} with browser JavaScript environment integration.  Pengine's knowledge base is extended with the programmer's Prolog modules.  Kernel module stores a common global part of A-~and T-boxes of applications, \emph{i.e.}, it is application database.  Other modules provide an application (C3) with services.

Authoring tool (B1) is used to provide document generation, and integration \cite{zont19}.  This tool enables document content and structure inference in the browser utilizing templates and data from other web pages and websites providing LOD\footnote{Abbreviation of Linked Open Data} \cite{lod} capabilities.  Programmer implements a view of an object as a web page with an extended version of RDFa.  The page is generated within any programming runtime, \emph{e.g.} PHP.  The authoring tool engine collects web page chunks and composes the document as an XHTML page.  This composed page can be stored in a KG as a set of triples and indexed in the text indexing module (C1), printed with the browser's printing engine, or converted to PDF.  The tool allows the user to change document content in WYSIWYG mode.  Its general structure is defined by changing the template on the server-side.

Text indexing engine (C1) provides a service for storing documents represented as KG triples, JSONs, and BLOBS of any format with corresponding text layers.  This module has two implementations.  One is built on the Sphinx Search engine, and the second is Elasticsearch.  Elasticsearch supports JSON as the only document format and is easily integrated with KG using JSON-LD for triple representation.  Sphinx Search is much faster in text and database records indexing and consumes less RAM as it is implemented in C++.  The choice of an engine depends on the primary document representation used in the application.

BLOB stored documents having no markup (PDFs, scanned documents) usually contain valuable data to be revealed for application use.  The documents could be report data or scientific paper content, DJVU files, or raster images.  Data recognition is implemented in the PDF/text document processing module (C2), where data processing workers query the text indexing module for BLOB documents and add recognized text and high-order structure layers.  All obtained layers are stored in the database of C1, and data of general interest are to be converted in triples of KG.

@@

The aim of this R\&D \cite{zont19, aicts2020} is to develop software support of the processes of NGS analysis with techniques and software for a visual representation of the computational process for so-called amplicon analysis so that domain specialists would be able to compose computational pipelines, which are executed on distributed heterogeneous computing resources (clouds).  Later, the visual models will be used to organize the computational process, data store, and particular NGS data retrieval services.

@@

All transformations were carried out with an object-oriented (OO) Logtalk programming environment representing modules that logically infer new data descriptions of the target objects.  Thanks to OO, the transformation has two levels of abstraction in its representation.  The higher one is a network of objects sending messages via their interfaces.  At the lower one is analyzing and synthesizing rules, generating new conclusions on object properties.

The PSM objects must have an order in their description as they represent a model of source code to be generated.  The order is organized with blocks of \verb|assert|ed statement and subobjects sequences referencing KG elements.  Blocks can be appended and pretended with new elements.  At the rendering sources stage, the three blocks are traversed in-depth.  This approach leverages the known \verb|llvmlite| source code generation.  Figure~\ref{fig:block-seq} shows the main idea behind PSM object synthesis.
\begin{figure}
  \centering
\begin{multicols}{2}
\begin{minted}[fontsize=\scriptsize]{logtalk}
:- object(class, specializes(code_block),
   imports([named])). % Category of named entities
:- public([classlist/1, methods/1, attributes/1]).
% . . . . . . . . . . . . . .
renderitem(Item, Result):-      % proceed with default
    ^^renderitem(Item, Result). % rendering
render(Result):-         % Source generator
    ^^render(Name),      % implemented in a category
    ( ::item(classlist(List)) ->
     % . . . . . . . . . . .
        [Name]) ),
    ( ::item(attributes(Attributes))->
     % . . . . . . . . . . .
        [DefAttrList]),
      Attributes::items(InstanceAttrs),
      findall(S, ( % initialize attributes
         % . . . . . . . . .
         ), AttrAssigns),
        root::unindent,
        AttrList=[ConstructorDef|AttrAssigns];
         % . . . . . . . . .
        AttrList=[ConstructorDef, Pass] ),
    ( ::item(methods(Methods))-> % If any ...
      Methods::render(MethodList);
      MethodList=[] ),
    lists::append(AttrList,MethodList,StringList),
    root::unindent, Result=[Signature|StringList].
:- end_object.
\end{minted}
      \includegraphics[width=0.7\linewidth]{code_block_class.pdf}
\end{multicols}
  \caption{Procedure of dataflow module source code model synthesis}
  \label{fig:block-seq}
\end{figure}
This example of the application of KG utilizes strongly the property of data representation flexibility for description of the software object model data as well as one central point of data store and access.  The disadvantageous properties of RDF, namely, the weak expressiveness with triples, was neglected by Prolog predicate and Logtalk object encapsulation.

@@

This project \cite{authoring} is aimed at the construction of an environment for document authoring out of user-editable template elements, database query results, LOD-compliant web pages, and other kind of a document content.  The target document is a web page where content is inferred while loaded by the browser.

All document elements are either \emph{context} structures or templates filled with other elements.  A context is an RDF hierarchical representation of a document data structure, providing data to be filled in a current rendered template.  The document is rendered from its main template.  All elements are SW resources.  Templates can contain JavaScript subroutines for implementing special services, for example, table of contents out of CSS \verb|class| labeled HTML headers of the generated document.

@@

<main lang="ru" resource="#annotation" typeof="oa:Annotation" id="main-doc-cnt">
<div property="oa:hasTarget" resource="#course-work-prog"></div>
<article property="oa:hasBody" typeof="foaf:Document curr:WorkingProgram"
         resource="#course-work-program" id="main-document">
  <div |\GB{taa:content}|="imei:title-page"></div>
  <div |\GB{taa:content}|="imei:neg-UMK"></div>
  <section id="TOC" class="break-after"> <h2>Table of Contents</h2>
    <div id="tableOfContents"></div>
  </section>
  <section id="course-description" resource="#description"
           property="schema:hasPart" typeof="schema:CreativeWork">
    <div property="schema:hasPart" resource="#purpose"
         typeof="dc:Text cnt:ContentAsText" >
      <div property="cnt:chars" datatype="xsd:string">
        <h2 property="dc:title" datatype="xsd:string">
           Aims and objectives of the discipline (module)</h2>
        <p>The aim of teaching the discipline ...</p>
      </div>
   </div>
  . . . . . . . .
\end{minted}
  \caption{Part of a template representation of an educational document}
  \label{fig:auth-code}
\end{figure}

For document markup (document data representation), the following vocabularies are used:
\begin{itemize}
  \item Friend-of-a-friend (\textbf{foaf}), agent information: individuals, legal entities, program agents;
  \item Provenance (\textbf{prov}), references between documents, data origins, proofs, \emph{etc.};
  \item Dublin Core (\textbf{dc}), web page annotation mark up;
  \item DBPedia resource (\textbf{dbr}), references to instant objects and classes;
  \item Schema.org (\textbf{schema}), Google, Yandex, Yahoo, \emph{etc}. searchable objects, structural elements;
  \item The Bibliographic Ontology (\textbf{bibo}), literature reference mark up.
\end{itemize}

@@

GIS represents spatial data in semantic-defined layers.  For each object of a GIS layer, one can associate a set of attribute values of auxiliary data.  The attribute set is the same for each object of the layer, regardless the attribute's value is defined or not.  Empty values are represented as ``\texttt{null}''s.  In the case of geological exploration, when many attributes are undefined, this approach leads to sparsely filled tables.  In this case, structure modification and data analysis algorithms are required to utilize additional data checking stages when using standard relation operations (SELECT, UPDATE, DELETE).

@@

\bibitem{tbl} T.~Berners-Lee, J.~Hendler, O.~Lassila.  The semantic web: A new form of web content that is meaningful to computers will unleash a revolution of new possibilities.  Scientific American, May 2001.

\bibitem{hogan} A.~Hogan, E.~Blomqvist, M.~Cochez, C.~D’Amato \emph{et al}. Knowledge Graphs, 2020. URL:\url{https://arxiv.org/abs/2003.02320v5} (access date: 12-Dec-2021)

\bibitem{zont19} E.~Cherkashin, A.~Shigarov, V.~Paramonov. Representation of MDA transformation with logical objects.  Procs. of International Multi-Conference on Engineering, Computer and Information Sciences (SIBIRCON), Novosibirsk, Russia, 2019, pp.~0913--0918 \doi{10.1109/SIBIRCON48586.2019.8958008}

\bibitem{cliopatria} J.~Wielemaker, W.~Beek, M.~Hildebrand, J.~Ossenbruggen, ClioPatria: A SWI-Prolog infrastructure for the Semantic Web, Semantic Web, 2016, volume 7(5). pp.~529--541 \doi{10.3233/SW"=150191}

\bibitem{swi} J.~Wielemaker, T.~Schrijvers, M.~Triska, T.~Lager. SWI-Prolog. Theory and Practice of Logic Programming, volume~12, No.~1-2, pp.~67--96, 2011, ISSN~1471-0684.

\bibitem{pengines} T.~Lager, J.~Wielemaker. Pengines: Web Logic Programming Made Easy.  Theory and Practice of Logic Programming, volume~14, No.~4-5, 2014, pp.~539--552. \doi{10.1017/S1471068414000192}

\bibitem{lod} C.~Bizer, T.~Heath, T.~Berners-Lee, Linked Data -- The Story So Far.  Int. J. Semantic Web Inf. Syst., volume~5, 2009, pp.~1--22. \doi{10.4018/jswis.2009081901}

\bibitem{aicts2020} E.~Cherkashin, A.~Shigarov. Construction techniques of Baikal microbiome research information-computational environment.  Procs. of 1st International Workshop on Advanced Information and Computation Technologies and Systems. Irkutsk, Russia, December 7-11, 2020. CEUR-WS.org, URL:\url{http://ceur-ws.org/Vol-2858/paper2.pdf} (access date: 12-Dec-2021)

\bibitem{gisviewer} E.~Cherkashin, O.~Lunina, L.~Demyanov, A.~Tsygankov. Web-GIS viewer for active faults data represented as a knowledge graph.  Procs. of 4th Scientific-practical Workshop Information Technologies: Algorithms, Models, Systems. Irkutsk, Russia, September 14, 2021. CEUR-WS.org, URL:\url{http://ceur-ws.org/Vol-2984/paper8.pdf} (access date: 12-Dec-2021)

\bibitem{lunina} O.~V.~Lunina.  The digital map of the Pliocene–Quaternary crustal faults in the Southern East Siberia and the adjacent Northern Mongolia.  Geodynamics \& Tectonophysics.  volume~7(3), 2016.  pp.~407-434. (in Russian) \doi{10.5800/GT-2016-7-3-0215}

\bibitem{afs} A.~A.~Gladkov, O.~V.~Lunina. Cartographic service ``Activetectonics''. URL:\url{http://activetectonics.ru/} (access date: 20-Sep-2021)

\bibitem{shigpaper} N.~O.~Dorodnykh, A.~Y.~Yurin, A.~O.~Shigarov. Conceptual Model Engineering for Industrial Safety Inspection Based on Spreadsheet Data Analysis.  In: Simian D., Stoica L. (eds) Modelling and Development of Intelligent Systems.  MDIS 2019.  Communications in Computer and Information Science, volume~1126, 2020, Springer, Cham. \doi{10.1007/978-3-030-39237-6_4}

@@
